\section{Key Difficulties and Innovations}

\subsection{Key Difficulties}

The focus lies in building an evaluation benchmark capable of precisely distinguishing different types of numerical biases and mapping human cognitive defects to model error patterns. The difficulties are twofold: First, effectively distinguishing information loss caused by the physical tokenizer level from schema confusion at the internal semantic representation level requires precise experimental design and deep interpretability analysis. Second, correcting deep-seated numerical biases via lightweight interventions without sacrificing the model's general language capabilities (avoiding "Catastrophic Forgetting")19.

\subsection{Innovations}

Fusion of Interdisciplinary Perspectives: Breaks the limitation of explaining errors solely from a CS perspective, systematically introducing "Conceptual Change Theory" and "Dual Process Theory" to analyze LLM numerical cognition, proposing a new perspective of "AI Conceptual Change."

Verification of Deep Mechanisms: Unlike previous black-box tests, this study attempts to verify the existence of "Version/Date Circuits" inhibiting logical reasoning inside the model using cutting-edge tools like Predictive Concept Decoders.

Human-Referenced Alignment Strategy: Proposes adapting teaching strategies used to overcome integer bias in human math education (e.g., emphasizing place value, using number line representations) to guide LLM alignment 20.

\section{Research Scheme and Implementation Plan}

Literature Review and Theory Preparation: Confirm direction, discuss with supervisor, collect literature on bias and LLM reasoning.

Experimental Design and Data Construction: Design comparative schemes. Script the "Human-AI Numerical Bias Alignment Dataset." Set up GPU environments for Llama 3/Qwen and API for GPT-4.

Implementation and Analysis: Execute large-scale testing. Analyze correlation with human data. Visualize internal states via probes. Conduct preliminary intervention experiments.

Writing and Mid-term Check: Organize data, draft the thesis, ensure logical rigor. Report progress regularly.

Subsequent Plan: Polish thesis, submit to conferences/journals if results are ideal, prepare for defense 22.
